Variation in use of East Asian Late Paleolithic weapons: A study of tip cross-sectional area of stemmed points from Korea

### 2022-10-09 analysis/paper/paper.qmd
Add text to the discussion section

abstract: |
 The introduction of blade technology, stemmed points, end scrapers, burins, denticulates, and finer grained materials led to the transition from the Early to Late Paleolithic in Korea. Stemmed points have been considered a representative tool that led this whole set of changes. In this research, we examine the function of stemmed points to understand the role that they played during the technological transition as well as throughout the Late Paleolithic. Our main questions are: What were stemmed points used for? How diverse were their functions? What are the temporal patterns in stemmed point functions? We measured tip cross-sectional areas (TCSA) to discriminate different functional classes of projectile points, for example, poisoned arrowheads or thrusting spear. We analyze TCSA with other variables including raw materials, weight, site and radiocarbon dates. Our results show that the stemmed points mostly functioned as javelins and thrusting spear tips, with smaller numbers as dart tips and arrowheads. TCSA values are depending on size and raw material types. We found different usage of stemmed points in different sites, which could indicate people used stemmed points in different ways depending on the environment. However, some sites show a wide range of TCSA values that represent multi-purpose usage of stemmed points. The temporal pattern of TCSA values is one of little change throughout the Late Paleolithic period. We conclude that stemmed points were mainly used as Javelin but they were multi-functional tools.
keywords: |
  Stemmed points; Korean Paleolithic; Tip Cross-Sectional Area
---

@@

## Choice of weapon-tip types

@@

>>>>

abstract: |
 The introduction of blade technology, stemmed points, end scrapers, burins, denticulates, and finer grained materials led to the transition from the Early to Late Paleolithic in Korea. Stemmed points have been considered a representative tool that led this whole set of changes. In this research, we examine the function of stemmed points to understand the role that they played during the technological transition as well as throughout the Late Paleolithic. Our main questions are: What were stemmed points used for? How diverse were their functions? What are the temporal patterns in stemmed point functions? We measured tip cross-sectional areas (TCSA) to discriminate different functional classes of projectile points, for example, poisoned arrowheads or thrusting spear. We analyze TCSA with other variables including raw materials, weight, site and radiocarbon dates. Our results show that the stemmed points mostly functioned as javelins and thrusting spear tips, with smaller numbers as dart tips and arrowheads. TCSA values are depending mostly on size rather than raw material types. We found different usage of stemmed points in different sites, which could indicate people used stemmed points in different ways depending on the environment. However, some sites show a wide range of TCSA values that represent multi-purpose usage of stemmed points. The temporal pattern of TCSA values is one of little change throughout the Late Paleolithic period. We conclude that stemmed points were mainly designed for Javelin and thrusting spear tips but they were multi-functional tools.
keywords: |
  Stemmed points; Korean Paleolithic; Tip Cross-Sectional Area
---

@@

## Two models for weapon-tip types choice

@@

To investigate the role that stemmed points played in the technological transition during the Korean Late Paleolithic, we applied TCSA metric to estimate artifact function by comparing the results to other archaeological and ethnographic cases. We asked three questions to achieve our research goal: What were stemmed points used for? How diverse were their functions? What are the temporal patterns in stemmed point functions? In our assumptions, human groups were capable of intentionally choosing and designing tools according to their socio-environmental circumstances. We came up with two models: (1) people created stemmed points and used them as a specific type of tool; or (2) they created stemmed points for multifunctional tools.

Our results indicate that javelins and thrusting spear tips are the best matches for stemmed points. There are a few stemmed points that can be used as dart tips and arrowheads, but none of them are suitable for poison arrowheads (@fig-tcsa-all-sp). We found stemmed points played a diverse role, by showing a wide range of TCSA values. Some sites, such as Nosanri and Sinbuk, used stemmed points as limited-function tools. Meanwhile, stemmed points were used in multiple ways at other sites, especially Suyanggae. In general, stemmed points played multiple roles in more sites (@fig-tcsa-sites). Through the Late Paleolithic period, TCSA values hardly changed (@fig-tcsa-radio-carbon). The chronological sequence of TCSA varies from region to region with no clear pattern. Therefore, we conclude that stemmed points were mainly designed for Javelin and thrusting spear tips but they were multi-functional tools.

We are aware of the concern that discriminating the function of small projectile points could be arbitrary work [@erlandson2014darts]. Because TCSA covers the critical elements of projectiles, flight and penetration dynamics (i.e. increased or decrease by shape of tip and cross section), we consider it a useful metric for discriminating between weapon tip functions [@hughes1998getting; @sitton2020tip; @lombard2021variation]. Stemmed points have been considered so far as long-distance hunting amateurs pursuing large bodies of prey such as a deer [@lee2019were; @lee2011study; @seong2008tanged; @lee2002new; @chang2013human]. TCSA enabled us to gain a more detailed understanding of the function of stemmed points.

TCSA can, however, be impacted by other factors such as raw materials and portability. -@eren2022plains explain that smaller TCSA could be the result of pursuing production economy and transport efficiency. We explore the relationship between TCSA and raw material types and weight in order to examine these factors. Weight plays an important role in portability and penetration of projectiles [@hughes1998getting]. According to our results, raw materials do not have a clear pattern, but weight is strongly correlated with raw materials (@fig-tcsa-raw-materials and @fig-tcsa-size). Different clusters of weight are matched with different types of weapon tips (@fig-artifact-size-cluster). Based on these analyses, we assume that people were able to produce the ideal size tools regardless of raw materials.
